Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects

### 2014-09-06 android-stacks.tex
Resolved language issues in intro.
And shortened it a bit.

Along the recent years, we have witnessed an astonishing increase in the number of
mobile applications. If on the one hand such applications extend phones capabilities 
far beyond of the basic calls and textual messages, on the other hand
they have to face an increasing number of threats to application robustness
 arising from: faults on underlying middleware and hardware (e.g., camera, sensors);
faults on third party services and libraries; compatibility issues ~\cite{McDon13}; 
 memory and battery restrictions; noisy external resources ~\cite{Zhang12} (e.g., wireless connections,GPS, bluetooth) among others.

Therefore, techniques for error detection and handling are not  an optional add-on but a 
fundamental part of such apps. The exception handling mechanism ~\cite{goodenough1975exception},
embedded in many mainstream programming languages, such as Java, C++ and C\#,
 is one of the most used techniques for detecting and recovering from such exceptional conditions. 
Android, the leading plaform for mobile applications~\cite{gartner} (thanks to its open-source 
model and hardware partners like Samsung, HTC, Motorola, Asus besides Nexus Google's
 own device)  reuses the embedded Java exception model to detect and handle 
 exceptional conditions.

Studies have shown, however, that the exception-related code is generally poorly understood and the
 less tested parts of the system~\cite{miller1997issues,Robil00,shah2010understanding, 
garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}.
As a consequence they may affect the system the other way around.
The exception-related code may favor the introduction of failures such as 
uncaught exceptions~\cite{jo2004uncaught, Zhang12} - 
which can lead to system crashes, making the system even less robust~\cite{coelho2011unveiling}.

@@

Until recent years, the availability of data, reporting such failures, was scarce. 
It comprised only few large scale projects (e.g. the Mozilla crash report dataset\footnote{https://code.google.com/p/promisedata/issues/detail?id=88}).
Recently we have observed a rapid increase in the number of open
source projects that make their version control systems as well as the issue tracking systems
publicly available in repository hosting sites such as GitHub and Google Code.
Developers and users of open-source projects uses such provided
 infrastrucute to share information (e.g., exception stack traces~\cite{bettenburg2008makes,schroter2010stack}) across 
the software development.

@@

This study performs a post mortem analysis of the exception stack traces embedded on issues
reported on Android projects hosted in GitHub and Google Code. 
The goal of this study was to investigate whether the reported exception stack traces
 can reveal common fragilities on the exception-related code.  
We call ``fragilities" characteristics on the exception-related code 
that favor the introduction of failures such as uncaught 
exceptions -  contributing to decrease the applicaiton robustness.
To guide this investigation we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective}.
Then, using a custom tool called ExceptionMiner,
 which we develop specifically for this study, we mine stack traces from the issues reported 
on 482 Android projects hosted in GitHub and 157 projects hosted in Google Code.
Overall 159,048 issues were analyzed and 6,005 stack traces were extracted from them.
The exception stack trace analysis was augmented with additional information
extracted from JDK, Android Plaform and the applications using byte code 
and source code analysis. Manual inspection was used to leverage the understanding 
of exception stack traces and support further discussions and insights.

Some outcomes consistently detected during this study were the following:

@@

       \item Approximately 52\% of the reported stack traces were associated to programming mistackes. 
And 27,71\% of all stack traces of reported a java.lang.NullPointerException the root cause.

       \item  4,4% of the reported exception stack traces ``undocumented runtime exceptions" signaled by third-party code. 
             (35 methods from Android platform and 44 from libraries).

     \item  Few cross-type exception wrappings, such as an OutOfMemoryError wrapped in a checked exception,
were detected on stack traces. Indicating that the purpose of Java's hybrid exception model may not have been adequately used.

   \item Undocumented checked exception signaled by a  JNI interface function defined on Android Platform. 
    Some flows reported a checked exception thown by a JNI interface function, and  such exception was not  
    part of the ``exception interface" of JNI function nor the methods on its invokation chain - it can lead to 
   uncaught exceptions that are difficult to debug.

@@

Such outcomes point to threats not only to the developement of robust Android apps, 
but also to the developement of any robust Java-based system. For instance, in the absence of the 
``exception specification" of  a third-party code, it is difficult or 
even enfeasible for the developer to protect the code against ``unforeseen" exceptions. 
Since, in such cases the client usually do not have access to the source code, such undocumented exceptions may 
remain uncaught and lead to system crashes. Moreover, the cross-type exception wrappings 
(e.g., OutOfMemoryError is wrapped in a checked Exception) 
may lead to unintended handling problems - e.g, trying to handle an instance of OutOfMemoryError ``hiden" in 
a checked exception may lead the program to an unpredictable state.

Hence, the study results are relevant to Android and 
Java developers who may underestimate how likely are the
 exceptions related to programming mistakes, and how 
 tricky it can be to prevent the uncaught exceptions from third-party code
 in the absence of ``exception interfaces". Moreover, results are relevant to 
designers of languages and tools to consider ways of reducing such fragilities 
in Android and Java environments.

The remaining of this paper is organized as follows. Section 2 presents a
background on Android Platform and Java exception model. 
Section 3 presents the study design. Section 4 reports the study findings. 
Section 5 provides further discussions and insights.
Section 6 presents the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 concludes the paper and outlines
directions for future work.

@@

\begin{figure*} \centering \includegraphics[scale=0.3]{studyprocess2.png}
\caption{Study overview.} \label{fig:overview} \end{figure*}

@@

\subsubsection{Android Apps in Google Code}
Google Code contains largely used open-source Android apps (e.g. K9Mail \footnote{K9Mail moved to Github but as a way of not loosing the project history 
it advises their users to report bugs on Google Code issue tracker.}), however, differently from GitHub, Google Code does not provide an API to access the information related
 to hosted projects \footnote{Google code used to provided a Web service to such repository, but it was deactivated in June 2013 what Google called a "clean-up action".}.
To overcome this limitation we needed to implement a Web Crawler (incorporated in ExceptionMiner tool described next) that navigated 
through the web interface of Google Code projects extracting all issues and issue comments and storing in a relational database for later analysis.

@@

In this paper we present an exploratory study in which we mined the stack 
traces embedded in all issues defined in 482 Android projects hosted in Github and 
157 projects hosted in Google Code. Overall it included 6,005 exception stack traces.
In this study, the information extracted from stack traces was used in combination 
with source code and bytecode analysis to identify common characteristics of exception 
stack traces and pinpoint fragilities on the exception-related code. 
We call ``fragilities" characteristics on the exception-related code that favor the introduction
of  uncaught exceptions and unintended handling - which can contribute to 
degrade the robustness of applications.
Some fragilities consistently detected in this study were: 
(i) an abundance of NullPointerExceptions - it was mentioned as the root exception in approximately 
50\% of the analyzed projects. 
(ii) undocumented runtime exceptions thrown by third party code,
(iii) undocumented checked exceptions thrown by JNI interface,
(iv) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions) 
revealing that Java hybrid exception model is not fully used according to its purpose.
As we can see, such fragilities can negatively affect not only the robustness of Android application 
but also the robustness of any Java-based application. 
Our results calls the attention for the need of tool and language support to help 
developers when dealing with such fragilities.

>>>>

In recent years, we have witnessed an astonishing increase in the number of
mobile applications. These applications extend phones capabilities 
far beyond basic calls and textual messages. Such applications, however,
have to face an increasing number of threats to application robustness
 arising from: failures in underlying middleware and hardware (e.g., camera, sensors);
failures in third party services and libraries; compatibility issues~\cite{McDon13}; 
 memory and battery restrictions; and noisy external resources ~\cite{Zhang12} (e.g., wireless connections,GPS, bluetooth).

Therefore, techniques for error detection and handling are not  an optional add-on but a 
fundamental part of such apps. The exception handling mechanism~\cite{goodenough1975exception},
embedded in many mainstream programming languages, such as Java, C++ and C\#,
 is one of the most used techniques for detecting and recovering from such exceptional conditions.
In this paper we will be concerned with exception handling in Android apps,
which reuse Java's exception handling model.

%Android, the leading plaform for mobile applications~\cite{gartner} (thanks to its open-source 
%model and hardware partners like Samsung, HTC, Motorola, Asus besides Nexus Google's
% own device)  reuses the embedded Java exception model to detect and handle 
% exceptional conditions.

Studies have shown that exception-related code is generally poorly understood and among the least tested parts of the system~\cite{miller1997issues,Robil00,shah2010understanding, 
garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}.
As a consequence they may inadvertently negatively affect the system: Exception-related code may introduce failures such as 
uncaught exceptions~\cite{jo2004uncaught, Zhang12} - 
which can lead to system crashes, making the system even less robust~\cite{coelho2011unveiling}.

@@

%% I am not so sure this data was not available. Let's omit this paragraph
%Until recently, the availability of data reporting such failures was scarce. 
%It comprised only few large scale projects (e.g. the Mozilla crash report dataset\footnote{https://code.google.com/p/promisedata/issues/detail?id=88}).
%Recently we have observed a rapid increase in the number of open
%source projects that make their version control systems as well as the issue tracking systems
%publicly available in repository hosting sites such as GitHub and Google Code.
%Developers and users of open-source projects uses such provided
% infrastrucute to share information (e.g., exception stack %traces~\cite{bettenburg2008makes,schroter2010stack}) across 
%the software development.

@@

This study performs a post mortem analysis of the exception stack traces 
included in issues
reported on Android projects hosted in GitHub and Google Code. 
The goal of this study is to investigate whether the reported exception stack traces
 can reveal common fragilities in the exception-related code.  
We call ``fragilities'' characteristics of the exception-related code 
that increase the likelihood of introducing failures such as uncaught 
exceptions -  leading to a decrease in application robustness.

To guide this investigation we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective}.
Then, using a custom tool called ExceptionMiner,
 which we develop specifically for this study, we mine stack traces from the issues reported 
on 482 Android projects hosted in GitHub and 157 projects hosted in Google Code.
Overall 159,048 issues were analyzed and 6,005 stack traces were extracted from them.
The exception stack trace analysis was augmented by means of byte code 
and source code analysis
 with additional exception-related information
extracted from the JDK, Android Platform and the application layer.
%Manual inspection was used to leverage the understanding 
%of exception stack traces and support further discussions and insights.

Outcomes of our study include:

% Some outcomes consistently detected during this study were the following:

@@

       \item Approximately 52\% of the reported stack traces can be attributed to programming mistakes. 
In particular, 27.71\% of all stack traces reported a \texttt{java.lang.NullPointerException} as its root cause.

       \item  4.4\% of the reported exception stack traces are ``undocumented runtime exceptions'' raised by third-party code. 
             (35 methods from the Android platform and 44 from libraries).

     \item  We detected several cross-type exception wrappings, such as an OutOfMemoryError wrapped in a checked exception,
in stack traces. This suggests that the purpose of Java's hybrid exception model comprising checked and unchecked exceptions may not have been adequately used.

   \item Undocumented checked exception signaled by a  JNI interface function defined on Android Platform. 
    Some flows reported a checked exception thrown by a JNI interface function, yet this exception was not  
    part of the ``exception interface'' of the JNI function nor the methods on its invocation chain. This can lead to 
   uncaught exceptions that are difficult to debug.

@@

Such outcomes point to threats not only to the developement of robust Android apps, 
but also to the developement of any robust Java-based system. For instance, in the absence of the 
``exception specification" of third-party code, it is difficult or 
even infeasible for the developer to protect the code against ``unforeseen'' exceptions. 
Since in such cases the client usually does not have access to the source code, such undocumented exceptions may 
remain uncaught and lead to system crashes. Moreover, the cross-type exception wrappings 
(e.g., OutOfMemoryError is wrapped in a checked Exception) 
may lead to unintended handling problems - e.g, trying to handle an instance of OutOfMemoryError ``hiden" in 
a checked exception may bring the program to an unpredictable state.

Hence, the study results are relevant to Android and 
Java developers who may underestimate how 
 exceptions are related to programming mistakes, and how 
 tricky it can be to prevent the uncaught exceptions from third-party code
 in the absence of ``exception interfaces". Moreover, the results are relevant to 
designers of languages and tools to consider ways of reducing such fragilities 
in Android and Java environments.

The remaining of this paper is organized as follows. 
Section~\ref{sec:back} presents a
background on Android Platform and Java exception model. 
Section~\ref{sec:study} presents the study design. 
Section~\ref{sec:exceptionminer} describes the ExceptionMiner tool we developed to conduct our study.
Section~\ref{sec:result} reports the findings of our study.
Section~\ref{sec:disc} provides a discussion of the wider implications of our results.
Section~\ref{sec:threats} presents the threats to validity associated to this study. 
Finally Section~\ref{sec:rele} describes related work, 
and Section~\ref{sec:conc} concludes the paper and outlines
directions for future work.

@@

\begin{figure*} \centering \includegraphics[scale=0.3]{studyoverview3.png}
\caption{Study overview.} \label{fig:overview} \end{figure*}

@@

\subsubsection{Android Apps in Google Code}
Google Code contains largely used open-source Android apps (e.g. K9Mail \footnote{K9Mail moved to Github but as a way of not loosing the project history 
it advises their users to report bugs on Google Code issue tracker: https://github.com/k9mail/k-9/wiki/LoggingErrors.}),
 however, differently from GitHub, Google Code does not provide an API to access the information related
 to hosted projects \footnote{Google code used to provided a Web service to such repository, but it was deactivated in June 2013 what Google called a "clean-up action".}.
To overcome this limitation we needed to implement a Web Crawler (incorporated in ExceptionMiner tool described next) that navigated 
through the web interface of Google Code projects extracting all issues and issue comments and storing in a relational database for later analysis.

@@

In this paper we present an exploratory study in which we mined the stack 
traces embedded in all issues defined in 482 Android projects hosted in Github and 
157 projects hosted in Google Code. Overall it included 6,005 exception stack traces.
In this study, the information extracted from stack traces was used in combination 
with source code and bytecode analysis to identify common characteristics of exception 
stack traces and pinpoint fragilities on the exception-related code. 
We call ``fragilities" characteristics on the exception-related code that favor the introduction
of  uncaught exceptions and unintended handling - which can contribute to 
degrade the robustness of applications.
Some fragilities consistently detected in this study were: 
(i) an abundance of NullPointerExceptions - it was mentioned as the root exception in approximately 
50\% of the analyzed projects. 
(ii) undocumented runtime exceptions thrown by third party code,
(iii) undocumented checked exceptions thrown by JNI interface,
(iv) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions) 
revealing that Java hybrid exception model is not fully used according to its purpose.
Such fragilities can negatively affect not only the robustness of Android application 
but also the robustness of any Java-based application. 
Our results calls the attention for the need of tool and language support to help 
developers when dealing with such fragilities.
